Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-06-11 paper1_jiturra.Rmd
comentarios method

Individual data from the "Social Networks and Social Resources"  module of the International Social Survey Programme (ISSP) from 2017 is employed [@issp_networks_2017]. This module includes information on a broader range of topics linked to social networks, attitudes toward economic inequality as well as demographic and socioeconomic background characteristics. The initial dataset includes 47,027 individual observations from 32 countries. However, one country (Slovenia) is excluded from the study because the main dependent variables are not available. Subsequently, after applying listwise case deletion, the analytical sample is composed of 32,529 individuals in 31 countries.

@@

For measuring economic inequality the Gini index (post-taxes and transfers) from the World Income Inequality Dataset (WID) is employed [@alvaredo_world_2022]. Additionally, two contextual variables are incorporated as controls in the multilevel models. Firstly, employing the Gross Domestic Product (GDP) available in the WID ensures consistency in the economic inequality estimates independently of economic property [@Schmidt-Catran2016; @finseraas_income_2009]. Secondly, to account for heterogeneity in institutional arrangements stemming from welfare schemes, a measure of the size of the welfare state is included [see @edlund_democratic_2015]. The measure is a standardized and rescaled indicator from 0 to 100 that combines (i) tax revenue as a percentage of GDP [@ilo_world_2022], (ii) welfare generosity as total governmental spending as a share of GDP [@ilo_world_2022], and (iii) the current level of redistribution [@solt_measuring_2020].

@@

Multilevel linear regression models are employed in all the analyses for accounting by the hierarchical structure of the data given that individuals are nested within countries. Therefore, the analysis begins with estimating the null model by declaring the nested structure using random intercept. This initial model assesses the intraclass correlation, revealing that 13.5% of the variance in redistributive preferences can be attributed to belonging to higher-level units. Subsequently, the micro-level models are estimated to determine the association between network homogeneity and the interaction with social class to test hypothesis 1. Following this, macro-level models are estimated by incorporating random intercepts alongside random slopes for network homogeneity and social class. In this set, a three-way cross-level interaction is estimated to determine if income inequality moderates the interaction between network homogeneity and social class to test hypothesis 2 [^1]. In the latter model, all the individual-level variables are group-mean centered (CWC) to mitigate possible collinearity problems [@hox_multilevel_2010]. Additionally, all the country-level factors have been standardized (z-scores) to ease comparability in the estimations [@hox_multilevel_2010]. For all the analyses, the package `lme4` for in the statistical package `R`  is employed [@bates_fitting_2015].

[^1]: Supplementary analyses employing alternative income inequality measures show that the results are robust when using the Inter-decile ratio (D9/D1) and the Top 10/Bottom 50 ratio. Additionally, countries were classified into low, middle-low, middle, middle-high, and high-income inequality groups based on quintiles according to the Gini index. Hence, country-fixed effects regressions are used to control for the cross-country differences and observed and unobserved societal characteristics. The results are consistent with the multilevel estimations.

@@

Regarding network segregation, Figure \@ref(fig:fig2) depicts the distribution of network homogeneity and social class across countries. In contrast to redistributive preferences, the distribution of network homogeneity is more scattered between countries and social classes. However, it is worth to mention at least two interesting findings. First, looking at the sample average (M = 0.36), we can notice that the variation between countries is low (SD = 0.08), with Thailand being the society with higher levels of network segregation (M = 0.43), and Iceland among lowest (M = 0.31). Second, it is depicted that the working class network homogeneity drives the pattern of homogeneity. In addition, the general pattern is that the service class is less segregated than the intermediate and working class. Besides, an interesting fact is that in some cases (e.g., Mexico), although the working class remains highly segregated (above 0.5), the intermediate and service classes are similar in their network homogeneity. As noticed, in most cases observed, the general pattern is that the working class is highly homogeneous in their networks, except in the Philippines, where the intermediate class shows a slightly higher average homogeneity. Overall, one of the most substantial findings is that the working class is more segregated than the services class.

@@

Moving to the bivariate macro relationships, Figure \@ref(fig:fig3) depicts the correlation between network homogeneity, redistributive preferences, and income inequality, our main societal characteristic of interest. First, according to @cohen_statistical_1988, for effect size criteria, we observe a medium positive and significant association between network homogeneity and redistributive preferences in Panel A (_r_ = 0.43, _p_ <.001). In concordance with the previously depicted distribution, we know that higher levels of network homogeneity are driven by the highly segregated networks of the working classes; what makes it logical that in countries where network homogeneity is high, this might also reflect greater social segregation among the working classes and consequently drive higher redistributive preferences. Second, Panel B depicts a positive but relatively weak and non significant association between income inequality and network homogeneity (_r_ = 0.28, _p_<.01). In other words, average homogeneity also tends to increase in societies with greater income gaps. However, the macro association only shows the general pattern between countries but does not allow us to look deeper into class differences according to the levels of inequality.

Interestingly, certain countries with higher levels of inequality and network homogeneity are societies that present class differences between the working class and the services class (e.g., Thailand or Mexico). By contrast, more egalitarian countries tend to be less homogeneous and have smaller class differences (e.g., Denmark or Sweden). As an additional exploratory analysis (see Fig. \@ref(fig:figapp1)), we have found that the association between income inequality and the gap in network homogeneity between the upper and lower classes tends to be moderate and statistically significant (_r_ = 0.3, _p_<.01). In other words, income inequality not only increase homogeneity overall but higher social distance between social classes as well.

@@

Model 1 integrates our measure of network homogeneity. Initially, as previously noted in the macro associations, individuals with more homogeneous networks tend to exhibit stronger redistributive preferences. This association persists in Model 2, where we control for network size. However, after including social class in Model 3, the relationship between homogeneity and redistributive preferences appears to reverse, with a reduction in statistical significance (_p_<0.10). Subsequently, Model 4 incorporates income, education in years and employment status. As expected, socioeconomic status is negatively associated with redistributive preferences regarding household economic resources and educational credentials. In addition, we find that people outside the labor market do not differ significantly from those who are economically active in their preferences. Finally, in Model 5, we see that by incorporating the interaction term between network homogeneity and social class, we observe that the working and intermediate class interaction terms are positive and statistically significant. Based on Model 5, Figure \@ref(fig:fig4) shows increased redistributive preferences in homogeneous working-class networks and the intermediate class. For the service class, the relationship is reversed, where more homogeneous networks for this social class decrease redistributive preferences. Altogether, these results provide evidence in favor of hypothesis 1, where the class divide in redistributive preferences becomes stronger for individuals with more homogeneous social networks.

@@

To facilitate the interpretation of the indicator, we have calculated a homogeneity indicator based on social distance:

>>>>

The primary individual data source is the "Social Networks and Social Resources" module of the International Social Survey Programme (ISSP)  [@issp_networks_2017]. The ISSP consists of a nationally representative probability sample of the adult population without substitution of each country. Each country applies a properly adapted questionnaire to guarantee the cross-cultural validity of the data and allow meaningful comparisons between countries. The questionnaire includes sections on social networks, attitudes toward economic inequality, and demographic and socioeconomic background characteristics. The complete full sample includes 47,027 observations in 32 countries. However, after inspecting the required information and applying listwise case deletion, the sample employed in the final analyses consists of 32,529 observations in 31 countries [^1].

[^1]: Slovenia is excluded from the study because the support for government redistribution measured by 'It is the responsibility of the government to reduce the differences in income between people with high incomes and those with low incomes' was not available in the dataset.

@@

To comparatively measure economic inequality, I include the Gini index (post-taxes and transfers) from the World Income Inequality Dataset (WID) [@alvaredo_world_2022].  In addition, two contextual variables are incorporated as controls in the multilevel regression. Firstly, employing the Gross Domestic Product (GDP) in constant 2017 USD (PPP) ensures consistency in the economic inequality estimates independently of economic property [@wiid_2023]. Secondly, to account for heterogeneity in institutional arrangements stemming from welfare schemes, a measure of the size of the welfare state is included [see @edlund_democratic_2015]. The measure is a standardized and rescaled indicator from 0 to 100 that combines (i) tax revenue as a percentage of GDP [@ilo_world_2022], (ii) welfare generosity as total governmental spending as a share of GDP [@ilo_world_2022], and (iii) the current level of redistribution [@solt_measuring_2020].

@@

Multilevel linear regression models are employed in all the analyses for accounting by the hierarchical structure of the data given that individuals are nested within countries. Therefore, the analysis begins with estimating the null model by declaring the nested structure using random intercept. This initial model assesses the intraclass correlation, revealing that 13.5% of the variance in redistributive preferences can be attributed to belonging to higher-level units. Subsequently, the micro-level models are estimated to determine the association between network homogeneity and the interaction with social class to test hypothesis 1. Following this, macro-level models are estimated by incorporating random intercepts alongside random slopes for network homogeneity and social class. This set estimates a three-way cross-level interaction to determine if income inequality moderates the interaction between network homogeneity and social class to test hypothesis 2 [^2]. In the latter model, all the individual-level variables are group-mean centered (CWC) to mitigate collinearity problems between the lower and higher-level predictors and spurious cross-level interaction coefficients [@aguinis_bestpractice_2013]. Additionally, all the country-level factors have been standardized (z-scores) to ease comparability in the estimations [@hox_multilevel_2010]. For this purpose, the package `lme4` [@bates_fitting_2015] for the statistical package `R` is employed in all the analyses.

[^2]: Supplementary analyses employing alternative income inequality measures show that the results are robust when using the Inter-decile ratio (D9/D1) and the Top 10/Bottom 50 ratio. Additionally, countries were classified into low, middle-low, middle, middle-high, and high-income inequality groups based on quintiles according to the Gini index. Hence, country-fixed effects regressions are used to control for the cross-country differences and observed and unobserved societal characteristics. The results are consistent with the multilevel estimations.

@@

Regarding network segregation, Figure \@ref(fig:fig2) depicts the distribution of network homogeneity and social class across countries. In contrast to redistributive preferences, the distribution of network homogeneity is more scattered between countries and social classes. However, it is worth to mention at least two interesting findings. First, looking at the sample average (M = 0.36), it can be noticed that the variation between countries is low (SD = 0.08), with Thailand being the society with higher levels of network segregation (M = 0.43), and Iceland among lowest (M = 0.31). Second, it is depicted that the working class network homogeneity drives the pattern of homogeneity. In addition, the general pattern is that the service class is less segregated than the intermediate and working class. Besides, an interesting fact is that in some cases (e.g., Mexico), although the working class remains highly segregated (above 0.5), the intermediate and service classes are similar in their network homogeneity. As noticed, in most cases observed, the general pattern is that the working class is highly homogeneous in their networks, except in the Philippines, where the intermediate class shows a slightly higher average homogeneity. Overall, one of the most substantial findings is that the working class is more segregated than the services class.

@@

Moving to the bivariate macro relationships, Figure \@ref(fig:fig3) depicts the correlation between network homogeneity, redistributive preferences, and income inequality, our main societal characteristic of interest. First,  Panel A depicts a medium positive and significant association between network homogeneity and redistributive preferences (_r_ = 0.43, _p_ <.001). In concordance with the previously depicted distribution, the higher levels of network homogeneity are driven by the highly segregated networks of the working classes; what makes it logical that in countries where network homogeneity is high, this might also reflect greater social segregation among the working classes and consequently drive higher redistributive preferences. Second, Panel B depicts a positive but relatively weak and non significant association between income inequality and network homogeneity (_r_ = 0.28, _p_<.01). In other words, average homogeneity also tends to increase in societies with greater income gaps. However, the macro association only shows the general pattern between countries but does not allow us to look deeper into class differences according to the levels of inequality.

Interestingly, certain countries with higher levels of inequality and network homogeneity are societies that present class differences between the working class and the services class (e.g., Thailand or Mexico). By contrast, more egalitarian countries tend to be less homogeneous and have smaller class differences (e.g., Denmark or Sweden). Besides, a supplementary analysis (see Fig. \@ref(fig:figapp1)) demonstrates that the association between income inequality and the gap in network homogeneity between the upper and lower classes tends to be moderate and statistically significant (_r_ = 0.3, _p_<.01). In other words, income inequality not only increase homogeneity overall but higher social distance between social classes as well.

@@

Model 1 integrates our measure of network homogeneity. Initially, as previously noted in the macro associations, individuals with more homogeneous networks tend to exhibit stronger redistributive preferences. This association is robust to network size in Model 2. However, after including social class in Model 3, the relationship between homogeneity and redistributive preferences appears to reverse, with a reduction in statistical significance (_p_> .05). Subsequently, Model 4 incorporates income, education in years and employment status. As expected, socioeconomic status is negatively associated with redistributive preferences regarding household economic resources and educational credentials. In addition,  people outside the labor market do not differ significantly from those who are economically active in their preferences. Finally,  incorporating the interaction term between network homogeneity and social class, Model 5 shows that the working and intermediate class interaction terms are positive and statistically significant. Based on Model 5, Figure \@ref(fig:fig4) shows increased redistributive preferences in homogeneous working-class networks and the intermediate class. For the service class, the relationship is reversed, where more homogeneous networks for this social class decrease redistributive preferences. Altogether, these results provide evidence in favor of hypothesis 1, where the class divide in redistributive preferences becomes stronger for individuals with more homogeneous social networks.

@@

To facilitate the interpretation of the indicator, the homogeneity indicator based on social distance is calculated:
